---
title: 'EDF 6938 Lecture 8: Basic Text Processing'
author: "A.C. Thomas -- FSSS"
date: "10/18/2015"
output: html_document
---

Until now we've gone in the forward direction from data frames onward: how we can work analyses given the data structure with some modest level of processing. Today we start to look backwards to assemble data sets from unstructured sources.

## Text Search Functions (In R)

These functions appear in **many** other languages, but their R counterparts are really useful to learn. First, we assume that everything we're working with is a string/character vector.

Here's one. Let's do a little work on it! [http://shakespeare.mit.edu/hamlet/full.html]

```{r}
hamlet <- readLines ("http://www.acthomas.ca/FSSS/data/hamlet-raw.txt")
hamlet[1:20]
```

OK. Here's what I want:

- Make this into a two-column matrix. Column 1: who's speaking the line. 2: The line.
- Process out the junk.
- Get ready to analyze text as spoken by each character.

## Exploratory Work

Let's introduce our first function: `grep`, which searches for patterns in text.

First, let's look for precise matches. This gives us the location of matches for the search string in the text.

```{r}
grep ("Hamlet", hamlet)
hamlet[grep ("Polonius", hamlet)]
```

Lots of them here. Where do they appear in the text?

```{r}
grep ("Hamlet", hamlet, value=TRUE)
```

Different than:

```{r}
grep ("HAMLET", hamlet, value=TRUE)[1:20]
```

Case sensitivity is important!

Some more searching:

```{r}
grep ("nunnery", hamlet)
grep ("nunnery", hamlet, value=TRUE)
grep ("sword", hamlet, value=TRUE)
grep ("sword|nunnery", hamlet, value=TRUE)
intersect (grep ("Swear", hamlet), grep ("sword", hamlet))
which (grepl ("nunnery", hamlet) & grepl (" to ", hamlet))

```

The task I want to accomplish: **Count the number of four-letter words used by each character in the play.**

We need to screen out those lines with all capitals -- they're the guide to who's speaking. But also useful! First we need a system for handling these kinds of matches.

## Regular Expressions

Enter [regular expressions](https://xkcd.com/208/) (or `regex`) -- syntax for generalizing strings of characters with wild cards.  

```{r}
test.string <- "Florida Gators"
grep("Florida", test.string)
grep("Floridian", test.string)
```

`grep` covers a search term for  a particular string. We're also going to bring in `gregexpr`, which *retrieves* specific matches for regular expressions, and `gsub`, which replaces any matches we see with another string.

```{r}
gsub("Gator", "Seminole", test.string)
test.string.2 <- "There ain't no Gator like a Florida Gator"
gsub("Gator", "Seminole", test.string.2)
```

`gsub()` is exceptionally useful for practicing regular expression matching, because it's immediately evident in single strings what happened and where. So we'll practice with this for now.

## Quantifiers in Regular Expressions

How often?
- `+` after a regexp means "1 or more times"
- `*` means "0 or more times"
- `?` means "0 or 1 times" (optional, once)
- `{n}` means "exactly n times"
- `{n,}` means "n or more times"
- `{n,m}` means "between n and m times (inclusive)"  

More general terms: the `[]` brackets let you specify a range of characters for a single character replacement:

```{r}
test.string
gsub("[rstuv]", "x", test.string)
gsub("[r-v]", "x", test.string)
test.string.2
gsub("[aeiou]", "y", test.string.2)

gsub("a", "found", "Aaaaaaaaaah")
gsub("g*", "found", "Aaaaaaaaaah")
gsub("a*", "found", "Aaaaaaaaaah")
gsub("a+", "found", "Aaaaaaaaaah")
gsub("a+?", "found", "Aaaaaaaaaah")
gsub("a{4}", "found", "Aaaaaaaaaah")
gsub("[A-z]{4}", "!", "Aaaaaaaaaah")
```

How about a capitalized word? What do the `*`, `+`, and `?` operators do?

```{r}
test.string.3 <- "There ain't no Gator like A Florida Gator"
gsub("[A-Z][a-z]*", "CapitalizedWord", test.string.3)
gsub("[A-Z][a-z]+", "CapitalizedWord", test.string.3)
gsub("[A-Z][a-z]+ [A-Z][a-z]+", "TwoCapWords", test.string.3)
gsub("[A-Z][a-z]* [A-Z][a-z]*", "TwoCapWords", test.string.3)
gsub("[A-Z][a-z]*? [A-Z][a-z]*?", "TwoCapWords", test.string.3)
```


## Another Data Source: Event Logs with Names Attached

```{r}
loadatbats <- load (url("http://www.acthomas.ca/FSSS/data/mlb2012atbats.RData"))
loadatbats
shortevents <- descriptions[1:100]
events1000 <- descriptions[1:1000]
```

Here's the challenge: Find all the names at the beginning of these statements to put together a full roster of players. Two more special terms:

- `^` means "match the beginning of the string"
- `$` means "match the end of the string"

We also want to extract the relevant part of the string. We can do this with `gsub()` (and I often do) but an easier way to do it is with `gregexpr()`. Let's test it out.

```{r}
test.string.3
capmatches <- gregexpr("[A-Z][a-z]*", test.string.3) ## "CapitalizedWord"
```

Well, that was ... helpful ... what now?

```{r}
regmatches (test.string.3, capmatches)
```

So let's do that now. We're going to need a little R-fu to make it work completely, but not too much.

We need an indicator for "name at the beginning of a string". How about

```{r}
namematch <- grep ("^[A-Z][a-z]+ [A-Z][a-z]", shortevents)
length(namematch)
```

Two missing. We can get them quickly:

```{r}
nonamematch <- grep ("^[A-Z][a-z]+ [A-Z][a-z]", shortevents, invert=TRUE, value=TRUE)
nonamematch
```

So let's modify it accordingly. How about all capitalized bits? We need to add special characters too.

```{r}
namematch <- grep ("^([A-Z][a-z'\\.])+", shortevents)
length(namematch)
```

Let's check it:

```{r}
capmatches <- gregexpr("^([A-Z][a-z'\\.])+", shortevents) 
regmatches (shortevents, capmatches)
```

Whoops. Let's fix it:

```{r}
capmatches <- gregexpr("^([A-Z]+[a-z'\\.]+)", shortevents) 
regmatches (shortevents, capmatches)
```

Add a space:

```{r}
capmatches <- gregexpr("^([A-Z]+[a-z'\\.]+ )+", shortevents) 
regmatches (shortevents, capmatches)
```

Wait... 

```{r}
shortevents[84]
regmatches (shortevents, capmatches)[[84]]
```

Add multiple possible spaces:

```{r}
capmatches <- gregexpr("^([A-Z]+[a-z'\\.]+ +)+", shortevents) 
regmatches (shortevents, capmatches)
regmatches (shortevents, capmatches)[[84]]
```

OK! Now let's go big.

```{r}
namematch <- grep ("^([A-Z]+[a-z'\\.]+ +)+", events1000)
length(namematch)
nonamematch <- grep ("^([A-Z]+[a-z'\\.]+ +)+", events1000, invert=TRUE, value=TRUE)
nonamematch
```

In case you couldn't tell -- this can take some time. But you might even find it fun. (I find it satisfying, like golf.)

(sigh) Hyphens and Camel-Case names:

```{r}
namematch <- grep ("^([A-Z][A-z'\\.-]+ +)+", events1000)
length(namematch)
nonamematch <- grep ("^([A-Z][A-z'\\.-]+ +)+", events1000, invert=TRUE, value=TRUE)
nonamematch
capmatches <- gregexpr("^([A-Z][A-z'\\.-]+ +)+", events1000) 
unlist (regmatches (events1000, capmatches))[101:200]
```

There's one more that I spotted on the manual check:

```{r}
regmatches (events1000, capmatches)[[492]]
events1000[492]
```

We can take care of this one either by removing all the "With" statements and starting again, or just removing "With" from our fits. 

```{r}
outcomes <- unlist (regmatches (events1000, capmatches))
outcomes[491:500]
gsub ("With ", "", outcomes[491:500])
```

Finishing touches.

```{r}
library(dplyr)
gsub ("With ", "", outcomes[491:500]) %>% gsub (" $", "", .)
```

## So, Back to the Bard

Let's have a look at that opening again. 

```{r}
hamlet[1:50]
```

Let's tag those lines that represent speakers. A couple of ways we can do it:

```{r}
speakerlines <- grep ("^[A-Z ]+$", hamlet[1:100], value=TRUE)
speakerlines
speakerlines.2 <- grep ("[a-z]", hamlet[1:100], invert=TRUE, value=TRUE)
speakerlines.2 ## not quite.
```

Still, we're learning something.

```{r}
hamlet <- hamlet[hamlet != ""]
hamlet <- grep("^(ACT |SCENE )", hamlet, invert=TRUE, value=TRUE)
```

Now: we want to identify the speakers.

```{r}
full.hamlet <- data.frame (speaker = "",
                           text = hamlet,
                           stringsAsFactors=FALSE)
speakerlines <- grepl ("^[A-Z ]+$", full.hamlet$text)
full.hamlet$speaker[speakerlines] <- full.hamlet$text[speakerlines]
  
head(full.hamlet, 20)
```

A little work here: if speaker is a blank, make it the speaker from the line before.

```{r}
for (thisrow in 2:nrow(full.hamlet)) full.hamlet$speaker[thisrow] <- if (full.hamlet$speaker[thisrow] == "") full.hamlet$speaker[thisrow-1] else full.hamlet$speaker[thisrow] 
head(full.hamlet, 20)
full.hamlet <- filter (full.hamlet, speaker != text)
head(full.hamlet, 20)
```

Spotted a couple more things:

```{r}
grep("^Enter ", full.hamlet$text, value=TRUE)
grep("^Exit ", full.hamlet$text, value=TRUE)
grep("^Exeunt ", full.hamlet$text, value=TRUE)
full.hamlet <- filter(full.hamlet, !grepl("^Enter ", text)) %>% filter (speaker != "", !grepl("^Exit ", text), !grepl("^Exeunt ", text))
head(full.hamlet, 20)
```

Let's get a reckoning now. How many lines does each person have?

```{r}
group_by (full.hamlet, speaker) %>% summarize (lines = n()) %>% arrange(desc(lines))
```

How many words does each person have? Let's count the spaces.

```{r}
full.hamlet <- mutate(full.hamlet,
                      spaces = sapply(regmatches(full.hamlet$text,gregexpr (" +", full.hamlet$text)), length) + 1)
head(full.hamlet)
group_by (full.hamlet, speaker) %>% summarize (lines = n(), words = sum(spaces)) %>% arrange(desc(words))
```

How many four letter words does each person have? We need an expression! We also need a "not" indicator, which is not at all confusing because it's the same symbol for "beginning"

```{r}
## pad it.
full.hamlet <- mutate (full.hamlet, text = paste0(" ",text," "))
test.set <- full.hamlet$text[1:20]
fourletterwords <- "[^A-z][A-z]{4}[^A-z]"
regmatches(test.set, gregexpr (fourletterwords, test.set))
full.hamlet <- mutate(full.hamlet,
                      fours = sapply(regmatches(full.hamlet$text,
                        gregexpr (fourletterwords, full.hamlet$text)), length))
head(full.hamlet)
```

Not quite! What went wrong?

```{r}
regmatches(test.set, gregexpr (fourletterwords, test.set))[[6]]
```

The first space got used in the first match, and not for the second.

```{r}
full.hamlet <- mutate (full.hamlet, mtext = gsub(" +","  ",text))
full.hamlet <- mutate(full.hamlet,
                      fours = sapply(regmatches(full.hamlet$mtext,
                        gregexpr (fourletterwords, full.hamlet$mtext)), length))
full.hamlet %>% select (-mtext) %>% head
```

```{r}
group_by (full.hamlet, speaker) %>% summarize (lines = n(), words = sum(spaces), fours = sum(fours)) %>% arrange(desc(fours)) %>% mutate (len = words/lines)
```


## In-Class Assignment

1. In how many lines does the word "die" appear in the text of Hamlet? 

2. In how many lines does a contraction with an apostrophe appear in the text of Hamlet?

3. How many times does a batter in baseball in 2012 strike out swinging? Strike out looking?

4. In what fraction of baseball events does a run score?

5. Find the number of contractions with apostrophes used by each character in Hamlet. What regular expression would you devise for this?






