---
title: "Lecture 2, Hour 1: Data Frames"
author: "A.C. Thomas -- FSSS"
date: "08/30/2015"
output: ioslides_presentation
---

## Data Frames from Scratch

A data frame in R is an object with 2-dimensional structure: by construction, each column is a variable with its own type and range.

- General case: Column types are unbounded, and can be variables of massive complexity.
- In this class: columns are each of some "atomic" type: numerical, character, factor, logical
- Pieces are joined together in order.


## What R Does

2 sorts of **objects**: **data** and **functions**

- **Data**:  things like 7, "seven", $7.000$, the matrix $\left[ \begin{array}{ccc} 7 & 7 & 7 \\ 7 & 7 & 7\end{array}\right]$

- **Functions**: things like $\log{}$, $+$ (two arguments), $<$ (two), $\mod{}$ (two), `mean` (one)

A function is a machine which turns input objects (**arguments**) into an output object (**return value**), possibly with **side effects**, according to a definite rule

## Data objects

All data is represented in binary format, by **bits** (TRUE/FALSE, YES/NO, 1/0)

- **Booleans/Logicals** Direct binary values: `TRUE` or `FALSE` in R
- **Integers**: whole numbers (positive, negative or zero), represented by a fixed-length block of bits
- **Characters** fixed-length blocks of bits, with special coding;
**strings** = sequences of characters
- **Floating point numbers**: a fraction (with a finite number of bits) times an exponent, like $1.87 \times {10}^{6}$, but in binary form
- **Missing or ill-defined values**: `NA`, `NaN`, etc.


## Constructing Data Frames

Here's a data frame constructed from scratch.

```{r}
some.states <- data.frame (state=c("Alabama", "Georgia", "Florida"),
                           abbrev = c("AL","GA","FL"),
                           pop = c("4779736", "9687673", "18801310"))
str(some.states)
summary(some.states)
```


## Loading in outside sources

```{r}
all.states <- read.csv ("NewStates.csv")
head(all.states, 2)
```


## Basic Operations: Row Subsets By Condition

Three methods:

```{r}
all.states[all.states$zone == "ESCent",]
subset (all.states, zone == "ESCent" & pop2005 > 3000)

library(dplyr)
filter (.data = all.states, zone == "ESCent", pop2005 > 3000)
```


## Basic Operations: Row Subsets By Position

```{r}
all.states[3:5,]
slice (all.states, 3:5)

all.states[13,]
slice (all.states, 13)
```

## Basic Operations: Column Subsets

Number of column:

```{r}
all.states[,2,drop=FALSE]
all.states[[2]]
```

Name of column:

```{r}
all.states[["Area"]]
all.states[,"Area",drop=FALSE]
```

Reduces this to a vector; removes geometry.

## Basic Operations: Column Subsets

dplyr:

```{r}
head(select (all.states, Area), 5)
```


## Basic Operations: Sorting

Let's take a subset of states.

```{r}
WSCent <- filter (all.states, zone == "WSCent")
WSCent
arrange (WSCent, pop2005)

WSCent$pop2005
order(WSCent$pop2005)
WSCent[order(WSCent$pop2005),]
slice(WSCent, rev(order(pop2005)))
```



## Basic Operations: Sorting

Let's take a subset of states.

```{r}
arrange (WSCent, desc(pop2005))
top_n (arrange(WSCent, desc(Area)), 2, pop2005)
```


## "Pipe" operator

Rather than nesting terms explicitly, the "pipe" from the R package `magrittr` helps to clean up syntax.

Default: make the item you "pipe" the first argument of the next function (parentheses optional, but helpful)

```{r}
sqrt(81)
81 %>% sqrt
65536 %>% sqrt()
```

If you don't want the first argument, use the placeholder `.` to mark the item you want replaced.

```{r}
pnorm(0.5, 0, 2)
2 %>% pnorm(0.5, 0, .)
```


## Group operations

Why this is so useful for our purposes: inline grouping by "key"(s). `group_by` function

```{r}
top_n(group_by(all.states, zone), 1, pop2005)
```

## Group operations

Combine this with the pipe and the order of operations is clear.

```{r}
all.states %>% group_by (zone) %>% top_n (1, pop2005)
```

## Group operations

Bottom up:

```{r}
all.states %>% group_by (zone) %>% top_n (1, desc(pop2005))
```


## Group operations

Combine this with the pipe and the order of operations is clear.

```{r}
all.states %>% group_by (zone) %>% arrange(Area)    ##
save.me.please <- all.states %>% group_by (zone) %>% top_n (1, desc(pop2005)) %>% ungroup %>% arrange(Area)    ##
```

```{r}
all.states %>% group_by(zone) %>% slice(10)
all.states %>% group_by(zone) %>% filter(pop2005 > median(pop2005))
all.states %>% group_by(zone) %>% filter(pop2005 >= median(pop2005))
all.states %>% group_by(zone) %>% mutate(pop.above.median = pop2005 - as.numeric(median(pop2005)))

```

## Group operations

Bottom up:

```{r}
all.states %>% group_by (zone) %>% top_n (1, desc(pop2005)) %>% ungroup %>% arrange(zone)
```


## Group summaries

Let's get some summary statistics on our group of US States.

```{r}
all.states %>% summarize (totalpop2005=sum(pop2005),
                          count=n(),
                          Area=sum(Area))
```

Grouping by zone:

```{r}
all.states %>% group_by (zone) %>%
  summarize (totalpop2005=sum(pop2005),
             count=n(),
             Area=sum(Area))
```


## Group summaries

Let's add some more categories for building a new data frame:

```{r}
all.states %>% group_by (zone) %>%
  summarize (totalpop2005=sum(pop2005),
             count=n(),
             Area=sum(Area),
             meanpop2005 = mean(pop2005),
             sdpop2005 = sd(pop2005) ##, CoD = sdpop2005/meanpop2005
             ) %>%
  mutate (CoD = sdpop2005/meanpop2005)
```



## Spread and Gather

This data frame is "tidy" if our experimental unit is "state". What about "state-year"? Recall

```{r}
WSCent
```

Use "gather"

```{r}
library(tidyr)
longtable <- gather (WSCent,  ## data frame for input.
                     year,    ## name for the "key"
                     population, ## name for the "value"
                     pop2005:pop1990)  ## columns to gather.
longtable
```

## Spread and Gather

```{r}
longtable <- longtable %>% mutate (year = substr(year, 4, 7))
plot(longtable$year, longtable$population)  
```

## Spread and Gather

```{r}
widetable <- longtable %>% spread (year, population)
widetable
```


## Setting up lagged data

We have lagged data implicitly in this table, but putting it in the long form would be useful too:

```{r}
longtable.2 <- longtable %>% group_by (State) %>% mutate (lagpop = lag(population))
head(longtable.2)
```

## Setting up lagged data

Whoops! Have to put it in order first because `lag()` doesn't care about numerical order, just vector order.

```{r}
longtable.2 <- longtable %>% group_by (State) %>% arrange (year) %>% mutate (lagpop = lag(population))
head(longtable.2)
```

## Let's add to it and check it.

```{r}
longtable.2 <- longtable %>% group_by (State) %>% arrange (year) %>% 
  mutate (lagpop = lag(population),
          growthrate = population/lagpop) %>% ungroup
slice(longtable.2, 5:8)
plot(growthrate ~ State, data = longtable.2)
```




